cardinalR: Generating interesting high-dimensional data structures

### 2025-09-07 paper-cardinalR.Rmd
Changes up to cubehole

The `gen_curvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) c"} else { "Figure \\@ref(fig:branch-proj) c"})`). The underlying geometry lies in the first two dimensions, while the remaining $(p-2)$ dimensions contain Gaussian noise.

Branch $1$ is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1X_1 + X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane. Branch $2$ is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane. Branches $3$ to $k$ are added iteratively. Each new branch $i$ begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X_{1}$:$X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure: $X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}$. Here, $s_i$ is a sampled scale factor (slope-like term) controlling curvature, and $(x_{\text{start}}, y_{\text{start}})$ is the starting point from the existing structure. This construction yields $k$ spatially connected, nonlinear branches with varying curvature. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

The `gen_orglinearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches embedded in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) d"} else { "Figure \\@ref(fig:branch-proj) d"})`). Each branch lies primarily within a distinct $2\text{-}D$ subspace, while the remaining $p - 2$ dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the $\binom{p}{2}$ possible $2\text{-}D$ combinations. If $k \leq \binom{p}{2}$, combinations are sampled without replacement. If $k > \binom{p}{2}$, additional pairs are sampled with replacement to reach $k$ total branches. Each selected pair $(X_{i1}, X_{i2})$ defines the $2\text{-}D$ plane for branch $i$.

For branch $i$, $n_i$ points are generated, where $\sum_{i=1}^k n_i = n$. The structure follows: $X_{i1} \sim U(0, 2)$, and $X_{i2} = -s_iX_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5)$, where $s_i$ is a scale factor controlling the slope. When $k \leq \binom{p}{2}$, $s_i = 1$. When sampling with replacement, $s_i$ is drawn from the set ${1, 1.5, 2, \dots, 8}$ in increments of $0.5$. For $p > 2$, the remaining dimensions contain independent Gaussian noise: $X_j \sim N(0, 0.1^2)$, for $j \notin \{i1, i2\}$.

@@

Let $\binom{p}{2}$ denote the number of unique $2\text{-}D$ subspace combinations. When $k \leq \binom{p}{2}$, $k$ distinct subspace pairs $(X_{i1}, X_{i2})$ are sampled without replacement. Otherwise, combinations are selected with replacement to reach $k$ total branches. For each branch $i = 1, \dots, k$, a random scale factor $s_i$ is used to vary the curvature; If $k \leq \binom{p}{2}$: $s_i = 1$, or If $k > \binom{p}{2}$: $s_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}$.

Each branch contains $n_i$ points such that $\sum_{i=1}^k n_i = n$, where the vector $(n_1, \dots, n_k)$ is randomly drawn using the helper function `gen_nsum()` to partition $n$. Within its assigned subspace $(X_{i1}, X_{i2})$, branch $i$ is defined by: $X_{i1} \sim U(0, 2)$, and $X_{i2} = -s_iX_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)$. This forms a smooth downward-opening parabola in the plane defined by $(X_{i1}, X_{i2})$, with the degree of curvature controlled by $s_i$. For $p > 2$, all remaining dimensions are independent Gaussian noise: $X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}$.

@@

To simulate a cone-shaped structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cone)"} else { "Figure \\@ref(fig:cone-proj)"})`) in arbitrary dimensions, we define a function `gen_cone(n, p, h, ratio)`, which creates a high-dimensional cone with options for a sharp or blunted apex, allowing for a dense concentration of points near the tip.

This function generates $n$ points in a $p\text{-}D$, where the last dimension ($X_p$) represents height along the cone's axis, and the remaining dimensions define a shrinking hyperspherical cross-section as one moves toward the tip. Points along the height axis are drawn from an exponential distribution to increase density near the tip. The $X_p = h_i \sim \text{Exp}(\lambda = 2/h)$ truncated at $h$. The effective radius at height $h_i$ decreases linearly from base to tip, controlled by a shape ratio parameter $r_i = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}})h_i/h, \quad \text{where } r_{\text{min}} = \text{ratio},\ r_{\text{max}} = 1$.

>>>>

The `gen_curvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) c"} else { "Figure \\@ref(fig:branch-proj) c"})`). The underlying geometry lies in the first two dimensions, while the remaining $(p-2)$ dimensions contain Gaussian noise.
Branch $1$ is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1X_1 + X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane. Branch $2$ is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane. Branches $3$ to $k$ are added iteratively. Each new branch $i$ begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ to ensure connectivity with earlier branches, spans a unit-length interval on $X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure: $X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, \; s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}$. Here, $s_i$ is a sampled scale factor (slope-like term) controlling curvature, and $(x_{\text{start}}, y_{\text{start}})$ is the starting point from the existing structure. This construction yields $k$ spatially connected, nonlinear branches with varying curvature. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

The `gen_orglinearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches embedded in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) d"} else { "Figure \\@ref(fig:branch-proj) d"})`). Each branch lies within a distinct $2\text{-}D$ subspace, while the remaining $p - 2$ dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the $\binom{p}{2}$ possible combinations of dimensions. If $k \leq \binom{p}{2}$, combinations are sampled without replacement. If $k > \binom{p}{2}$, additional pairs are sampled with replacement to reach $k$ total branches. Each selected pair $(i_1, i_2)$ defines the $2\text{-}D$ plane for branch $i$.

For branch $i$, $n_i$ points are generated, where $\sum_{i=1}^k n_i = n$. The structure follows: $X_{i_1} \sim U(0, 2)$, and $X_{i_2} = -s_iX_{i_1} + \epsilon, \quad \epsilon \sim U(0, 0.5)$, where $s_i$ is a scale factor controlling the slope. When $k \leq \binom{p}{2}$, $s_i = 1$. When sampling with replacement, $s_i$ is drawn from the set ${1, 1.5, 2, \dots, 8}$ in increments of $0.5$. For $p > 2$, the remaining dimensions contain independent Gaussian noise: $X_j \sim N(0, 0.1^2)$, for $j \notin \{i_1, i_2\}$.

@@

Let $\binom{p}{2}$ denote the number of unique $2\text{-}D$ subspace combinations. When $k \leq \binom{p}{2}$, $k$ distinct subspace pairs $(i_1, i_2)$ are sampled without replacement. Otherwise, combinations are selected with replacement to reach $k$ total branches. For each branch $i = 1, \dots, k$, a random scale factor $s_i$ is used to vary the curvature; If $k \leq \binom{p}{2}$: $s_i = 1$, or If $k > \binom{p}{2}$: $s_i \in \{1, 1.5, \dots, 8\}$.

Each branch contains $n_i$ points such that $\sum_{i=1}^k n_i = n$, where the vector $(n_1, \dots, n_k)$ is randomly drawn using the helper function `gen_nsum()` to partition $n$. Within its assigned subspace, branch $i$ is defined by: $X_{i_1} \sim U(0, 2)$, and $X_{i_2} = -s_iX_{i_1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)$. This forms a smooth downward-opening parabola in the plane defined by $(X_{i_1}, X_{i_2})$, with the degree of curvature controlled by $s_i$. For $p > 2$, all remaining dimensions are independent Gaussian noise: $X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i_1, i_2\}$.

@@

To simulate a cone-shaped structure in arbitrary dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cone)"} else { "Figure \\@ref(fig:cone-proj)"})`), we define a function `gen_cone(n, p, h, ratio)`, which creates a high-dimensional cone with options for a sharp or blunted apex, allowing for a dense concentration of points near the tip.

This function generates $n$ points in $p\text{-}D$, where the last dimension ($p$) represents height along the cone's axis, and the remaining dimensions define a shrinking hyperspherical cross-section as one moves toward the tip. Points along the height axis, $p$, are drawn from an exponential distribution to increase density near the tip. That is, $X_p = h_i \sim \text{Exp}(\lambda = 2/h)$ truncated at $h$. The effective radius at height $h_i$ decreases linearly from base to tip, controlled by a shape ratio parameter $r_i = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}})h_i/h, \quad \text{where } r_{\text{min}} = \text{ratio},\ r_{\text{max}} = 1$.
